To Airbnb?: A look at Long vs Shorm Term Leasing

### 2018-10-31 papers/full_paper.Rmd
Paper updates to Owner characteristics in data.

---
title: "To Airbnb?: A Question of Revenue"
#subtitle: "(Andy Krause and Gideon Aschwanden)Rev"
#author: "Greenfield Advisors and University of Melbourne"
date: "Revised: July 22nd, 2018"
output:
  pdf_document: default
  html_document: default
header-includes: 
   - \usepackage{setspace}\doublespacing
   - \usepackage{caption}
   - \captionsetup[table]{skip=10pt}
geometry: margin=3.3cm
bibliography: abb.bib
editor_options: 
  chunk_output_type: console
---

@@

The growth of Airbnb and other short-term rental platforms have presented owners of residential properties with a choice of leasing strategy: traditional long-term rental or a short-term approach ("Airbnb-ing").  In this paper we identify those situations -- location, structure type and property characteristics -- that lead to the highest likelihood of favoring a short-term strategy over a long term one. We also find that hosting policies matter, suggesting that even the right property needs the right owner(s) to make short-term rental the more profitable approach.

@@

With the rise of Airbnb and other self-hosting accommodation platforms^[Flipkey.com, VRBO.com, Homeaway.com, Tripping.com, Stayz.com.au as examples.], it is easier than ever for property owners and investors to consider short-term rental (STR) as a potential leasing strategy for a revenue generating residential property. Should they though? Or rather, is it more profitable to do so when compared to traditional long-term rental (LTR)? And further, are certain types of properties in specific locations more likely to show a preference for a short-term leasing strategy over long-term one?

To address these questions and to provide some guidance to potential and current investors and owners, we examine the short- and long-term residential rental markets in Melbourne, Australia. Melbourne has a thriving short-term rental market with over 20,000 total properties listed on the Airbnb portal, the source of our short-term market data.

We adopt a scenario-based approach that compares observed annual revenues for short-term rental (STR) properties with an estimate of their expected long-term rental revenues. Our results show that the feasibility of a greater return through a short-term rental -- what we have termed ``Short-Term Preference'' -- is dependent on property type, product, location and host policies.  In short, the answer to whether or not a owner or investor should opt for a short-term leasing strategy is ''it depends''.

@@

We approach answering the question of whether it is more profitable to pursue a short- or long-term leasing strategy by creating and assessing a scenario likely to be faced by a recent purchaser of a resdiential investment property.  More specifically, we examine the following case:

* An investor purchases a property in Melbourne, Australia on September 1, 2015.
* They are looking to maximize the net revenue over the 12-month period from September 1, 2015 to August 31, 2016.
* The choices are 1) traditional year long rental (long-term), or: 2) short-term, nightly Airbnb-type rental.
* Only direct monetary revenues and costs are considered, others such as liquidity and inconvenience are ignored.
* As most long-term rentals in Melbourne use a leasing agency, we will assume use of an agency.

We will not, a priori, specify a location or property type for the scenario as we hypothesize that the preference of a short-term rental over a long-term one will vary by the location and property characteristics. The extent to which location, property type and property characteristics impact the probability of short-term preference are examined in the results section.

@@

As our scenario compares a hypothetical and particular owner over the course of a single  year the tax $T$ and $F$ costs are identical between the two leasing strategy and therefore cancel out. General maintenance $M$ expenses may differ between the two strategies, and an argument could be made for either to be higher. Long-term renters spend more time in the property over the course of the year and bring their own appliances^[White goods are usually not supplied in the Australian rental market.] and furniture which may damage the home. They are also more likely to affix items to the structure.  Conversely, short-term rentals are usually furnished and the occupants have less incentive not to be evicted as their damage is not often uncovered until after they have left.^[Note that Airbnb's rating policy may help protect hosts as bad guests can be flagged on the platform.]  Both arguments are meritous and we were unable to find any evidence to support maintenance costs being higher in short- or long-term rentals. As a result we have made the assumption that the maintenance costs will be identical, and they too cancel out.

@@

Janitorial costs, $J$, cover the costs to 'turn-over' the property between occupants.  For a short-term leasing strategy, these costs can be experienced every day or two, depending on the minimum length of stay and the total number of occupancies over the course of the year. Our data for short-term rentals comes from the Airbnb platform.  In addition to the nightly rate that an occupant pays to the owner there is also a separate 'cleaning fee' that is attached, as is a booking fee to Airbnb.^[This fee could be considered analogous to the leasing fees $L$ that long-term rentals pay.] Some hosts add additional 'linen' fees to cover costs of laundering.  While it may be advantageous for hosts in competitive markets to underquote their cleaning fees to keep occupancies up, we have no evidence to suggest that owners are exhibiting such behavior. As a result of janitorial fees being a separate line item in the STR data (paid separately and in addition to the nightly rate) they can also be ignored in the comparative analysis below.

@@

Throughout this analysis we will refer to any situations where the short-term net revenue exceeds the long-term net revenue as 'Short-Term Preferable' or a situation of 'Short-Term Preference'.  For the purposes of this study we consider this a binary position leading to a binary decision of which leasing strategy to follow.

@@

Data for this study were gathered from two different sources, one for the short-term rental market and one for the long-term rental market.  Short-term rental market observations are from the Airbnb online portal. Airbnb is the largest short-term rental hub in the Melbourne region. As opposed to similar sites such as VRBO.com and Homeaway.com which focus on vacation homes in tourist dominated areas such as beach towns and ski resorts, Airbnb offers considerable coverage in urban areas and, therefore, directly competes to a greater degree with more traditional long-term rentals than the other sites. The data on the Airbnb market in Melbourne have been purchased from www.airdna.co, a data provider specializing in Airbnb data collection and analysis. The short-term data include two types of observations, property-level and nightly-per-property. Property-level data consists of a single observation for each property that is or has been listed on the Airbnb portal over the October 2014 to August 2016 period.^[This time period represents the period in which AirDNA was engaged in data collection in Melbourne.]  The nightly data offer one observation for each property for each night of the year during its time on the portal.  The nightly data indicate the advertised price, the occupancy status and a unique reservation identification.

@@

Data on the long-term rental market was provided by Australia Property Monitors (APM), a Fairfax Group company. In raw form this dataset includes an observation for each rental property listing for each time a detail of the listing changed. For example, if a property was listed for rent for one month and the price was dropped once each week there would be four observations in the dataset. These four observations are grouped by an event-specific identification number. This number is specific to the listing event, not the property. In each case the last observation signifies the final price of the rental contract. We separate this data into two sets for efficiency reasons. The first data set includes all information about the property, physical and economic (date and amount of rental contract), while the second contains a property ID, a rental event ID, the list price and the date of the change. As such, the first dataset is a set of market observations with associated characteristics and the second is a listing history of each rental event. The long-term lease data cover the period from January 2014 to December 2015.

Unfortunately, there are no common identifiers between the two datasets as the Airbnb data does not include addresses. Both sets of data do include latitude and longitude, but in a dense urban area with many multiple-family dwellings it is impossible to match apartments based on two-dimensional coordinates. As a result of this, we are unable to directly match observations from the two datasets. However, the number of properties that rented long-term for the 2014 to 2015 period and then switched to Airbnb for the 2015 to 2016 period are likely very small and this inability to directly match properties does not greatly impact our modeling strategy.

@@

As long-term rentals do not offer Private Room or Shared Room options^[Our long-term rental dataset does not include rooming houses, purpose built student accommodations and other types of rentals which may offer private or shared room options] and our purpose here is a comparison of long- and short-term returns, we remove all short-term properties that do not lease the entire home or apartment. Unfortunately, this filter does remove about 46% of the short-term data, however, given the research question it is unavoidable.

@@

>>>>

---
title: "To Airbnb?: A look at Long vs Shorm Term Leasing"
subtitle: "(Andy Krause and Gideon Aschwanden)Rev"
author: "Greenfield Advisors and University of Melbourne"
date: "Revised: October 16th, 2018"
output:
  pdf_document: default
  html_document: default
header-includes: 
   - \usepackage{setspace}\doublespacing
   - \usepackage{caption}
   - \captionsetup[table]{skip=10pt}
geometry: margin=3.3cm
bibliography: abb.bib
editor_options: 
  chunk_output_type: console
---

@@

The growth of Airbnb and other short-term rental platforms have presented absentee owners of residential properties with a choice of leasing strategy: traditional long-term rental or a short-term approach ("Airbnb-ing").  In this paper we identify those situations -- location, structure type and property characteristics -- that lead to the highest likelihood of favoring a short-term strategy over a long term one. We also find that hosting policies matter, suggesting that even the right property may need the right owner(s) or strategy to make short-term rental the more profitable approach.

@@

With the rise of Airbnb and other self-hosting accommodation platforms^[Flipkey.com, VRBO.com, Homeaway.com, Tripping.com, Stayz.com.au as examples.], it is easier than ever for property owners and investors to consider short-term rental (STR) as a potential leasing strategy for a revenue generating residential property. Should they though? Or rather, is it more profitable to do so when compared to traditional long-term rental (LTR)? Are certain types of properties in specific locations more likely to show a preference for a short-term leasing strategy over long-term one? Do host policies matter?

To address these questions and to provide some guidance to potential and current investors and owners, we examine the short- and long-term residential rental markets in Melbourne, Australia. Melbourne has a thriving short-term rental market with over 20,000 total properties listed on the Airbnb portal, the source of our short-term market data. To avoid confounding market dynamics caused by traditional vacation rentals (beach, ski and other recreation) we limit our study area to metropolitan area and suburban/rural areas lacking these recreational amenities. Additionally, for any individual property the choice between short and long term leasing may be influenced by individual financial situations (tax and mortgage financing) as well as non-financial concerns such as time availability and personal attachement to the property.  While we analyze individual properties, we acknowledge that there are many unobserved condiseration and, therefore, we generalize our findings to market segments, locations and host policy decisions.

We adopt a scenario-based approach that compares observed annual revenues for short-term rental (STR) properties with an estimate of their expected long-term rental revenues, netting out costs that can be estimated or assumed with some generality. We handle potential undercounting of costs through sensitivity tests that follow our initial assessment.

Our results show that the feasibility of a greater revenues through a short-term rental -- what we have termed ``Short-Term Preference'' -- is dependent on property type, product, location and host policies.  In short, the answer to whether or not a owner or investor should opt for a short-term leasing strategy is ''it depends''.

@@

We approach answering the question of whether it is more profitable to pursue a short- or long-term leasing strategy by creating and assessing a scenario likely to be faced by a recent purchaser of a residential investment property.  More specifically, we examine the following case:

* An investor purchases a property in Melbourne, Australia on September 1, 2015.
* They are looking to maximize the net revenue over the 12-month period from September 1, 2015 to August 31, 2016.
* They do not intend to inhabit the home themselves. 
* The choices are 1) traditional year long rental (long-term), or: 2) short-term, nightly Airbnb-type rental.
* Only direct monetary revenues and costs are considered, others such as liquidity and inconvenience are ignored as these concerns are influenced by owner situations, not property characteristics or hosting policies.  
* As most long-term rentals in Melbourne use a leasing agency, we will assume use of an agency.

@@

As our scenario compares a hypothetical and particular owner over the course of a single year the tax $T$ and $F$ costs are identical between the two leasing strategies and therefore cancel out. General maintenance $M$ expenses may differ between the two strategies, and an argument could be made for either to be higher. Long-term renters spend more time in the property over the course of the year and bring their own appliances^[White goods are usually not supplied in the Australian rental market.] and furniture which may damage the home. They are also more likely to affix items to the structure.  Conversely, short-term rentals are usually furnished and the occupants have less incentive not to be evicted as their damage is not often uncovered until after they have left.^[Note that Airbnb's rating policy may help protect hosts as bad guests can be flagged on the platform.]  Both arguments are meritous and we were unable to find any evidence to support maintenance costs being higher in short- or long-term rentals. As a result we have made the assumption that the maintenance costs will be identical, and they too cancel out.

@@

Janitorial costs, $J$, cover the costs to 'turn-over' the property between occupants.  For a short-term leasing strategy, these costs can be experienced every day or two, depending on the minimum length of stay and the total number of occupancies over the course of the year. Our data for short-term rentals comes from the Airbnb platform. On Airbnb, in addition to the nightly rate that an occupant pays to the owner there is also a separate 'cleaning fee' that is attached, as is a booking fee to Airbnb.^[This fee could be considered analogous to the leasing fees $L$ that long-term rentals pay.] Some hosts add additional 'linen' fees to cover costs of laundering.  While it may be advantageous for hosts in competitive markets to underquote their cleaning fees to keep occupancies up, we have no evidence to suggest that owners are exhibiting such behavior. As a result of janitorial fees being a separate line item in the Airbnb data (paid separately and in addition to the nightly rate) they can also be ignored in the comparative analysis below.

@@

Throughout this analysis we will refer to any situations where the short-term net revenue exceeds the long-term net revenue as 'Short-Term Preferable' or a situation of 'Short-Term Preference' (STP).  For the purposes of this study we consider this a binary position leading to a binary decision of which leasing strategy to follow.

@@

Data for this study were gathered from two different sources, one for the short-term rental market and one for the long-term rental market.  Short-term rental market observations are from the Airbnb online portal. Airbnb is the largest short-term rental hub in the Melbourne region. As opposed to similar sites such as VRBO.com and Homeaway.com which focus on vacation homes in tourist dominated areas such as beach towns and ski resorts, Airbnb offers considerable coverage in urban areas and, therefore, directly competes to a greater degree with more traditional long-term rentals than the other sites.^[In fact, our choice of study area was done so to specifically minimize the number of potential traditional vacation homes in our sample.] The data on the Airbnb market in Melbourne have been purchased from www.airdna.co, a data provider specializing in Airbnb data collection and analysis. The short-term data include two types of observations, property-level and nightly-per-property. Property-level data consists of a single observation for each property that is or has been listed on the Airbnb portal over the October 2014 to August 2016 period.^[This time period represents the period in which AirDNA was engaged in data collection in Melbourne.]  The nightly data offer one observation for each property for each night of the year during its time on the portal.  The nightly data indicate the advertised price, the occupancy status and a unique reservation identification.

@@

Data on the long-term rental market was provided by Australia Property Monitors (APM), a Fairfax Group company. In raw form this dataset includes an observation for each rental property listing for each time a detail of the listing changed. For example, if a property was listed for rent for one month and the price was dropped once each week there would be four observations in the dataset. These four observations are grouped by an event-specific identification number. This number is specific to the listing event, not the property. In each case the last observation signifies the final price of the rental contract. The long-term lease data cover the period from January 2014 to December 2015.

Unfortunately, there are no common identifiers between the two datasets as the Airbnb data does not include addresses. Both sets of data do include latitude and longitude, however the accuracy of the Airbnb coordinates is suspect as these are often purposefully jittered to protect property anonymity and, as a result, in a dense urban area with many multiple-family dwellings it is impossible to match apartments based on imprecise two-dimensional coordinates. As a result of this, we are unable to directly match observations from the two datasets.

@@

As long-term rentals do not offer Private Room or Shared Room options^[Our long-term rental dataset does not include rooming houses, purpose built student accommodations and other types of rentals which may offer private or shared room options] and our purpose here is a comparison of long- and short-term revenues, we remove all short-term properties that do not lease the entire home or apartment. Unfortunately, this filter does remove about 46% of the short-term data, however, given the research question it is unavoidable.

@@

### Owner Types
